---
title: "Final Project"
author: 
    name: "Monique Hernandez"
    email: mahernandez@ucsb.edu
    affiliation: 
    - University of California, Santa Barbara
    - Masters of Environmental Data Science, 2027
    - B.S. in Environmental Studies, 2025
date: "Augustt 28, 2026"
toc: true
format:
  html:
    theme: "pulse"
    other-links:
      - text: EDS 214 Course Website
        href: https://eds-214-reproducibility-m26.github.io/course-materials/final-project/final-project.html
      - text: Source Code
        href: https://github.com/moniqueallyson/MAH214final
execute:
  message: false
  warning: false
---


# Background
The final project is the reproduction of figure 3 from the article by Douglas A. Schafer titled, "Effects of hurricane disturbance on stream water concentrations and fluxes in eight tropical forest watersheds of the Luquillo Experimental Forest, Puerto Rico". The final output is a graph depicting a 9 week average from the years between 1985 to 1995, from four locations in Puerto Rico: Bisley Quebrada 1-3 (BQ1-3) and Puente Roto Mameyes. In the article, the results show that Nitrate, Ammonium, and Potassium have increased, while Magnesium and Calcium did not that hurricanes can have an effect on streamwater. The hurricane disturbance had a different effect on different chemicals. The most disturbed watersheds were at the sites BQ1-3.

# Data

The data was downloaded on the website for EDI dAta Portal from William H. McDowell at the International Institute of Tropical Forestry(IITF), USDA Forest Service.

```{r}
library(tidyverse)
figure3 <- read_csv(
  "/Users/master/Downloads/Bren/MEDS Summer/eds214/MAH214final/output/clean_data.csv"
)
```


# Methods
The first steps to reproducing the figure from Schaefer's article was to sort through the raw data by cleaning it using select and filter functions. 
Sample_Date, Sample_ID, `NH4-N`, `NO3-N`, Ca, Mg, K
and filtering through the correct time period which was from 1988 to the end of 1994.


# Results

the v-line shown going down the graph represents the hurricane hugo formation.

```{r, message=FALSE}

ggplot(
  data = figure3, # reading the clean data csv in output
  mapping = aes(
    # gives the x-axis, y-axis, and the line by site location for each chemical ion
    x = Sample_Date,
    y = concentration,
    linetype = site
  )
) +
  ggtitle(
    "Hurricane effect on stream chemistry" # title from the article to give reference on what the figure is about
  ) +
  geom_line() +
  geom_vline(
    # to show reference line as shown in original figure which depicts the start of the distrubance
    xintercept = ymd("1989-09-18"), # hurricane hugo date formation
    linetype = "dashed", # dash line for the separation
    linewidth = 0.4
  ) +
  labs(
    x = "Years",
    y = NULL
  ) +
  facet_wrap(
    ~ion,
    ncol = 1,
    scales = "free",
    strip.position = "left",
    nrow = 5
  ) + # frees up the scaling for each ion since the units are differnet and so are the concentration levels
  theme_bw() + # black and white to match the original
  theme(
    legend.title = element_blank() # original did not have a legend title
  ) +
  theme(
    panel.grid.major = element_blank(), # remove major grid lines for cleaner look
    panel.grid.minor = element_blank() # remove minor grid lines for sleekness
  )
print(ggplot)
```

