---
title: "Final Project"
author: 
    name: "Monique Hernandez"
    email: mahernandez@ucsb.edu
    affiliation: 
    - University of California, Santa Barbara
    - Masters of Environmental Data Science, 2027
    - B.S. in Environmental Studies, 2025
date: "August 28, 2026"
toc: true
format:
  html:
    theme: "pulse"
    other-links:
      - text: EDS 214 Course Website
        href: https://eds-214-reproducibility-m26.github.io/course-materials/final-project/final-project.html
      - text: Source Code
        href: https://github.com/moniqueallyson/MAH214final
execute:
  message: false
  warning: false
---

# Background
The final project is the reproduction of figure 3 from the article by Douglas A. Schafer titled, "Effects of hurricane disturbance on stream water concentrations and fluxes in eight tropical forest watersheds of the Luquillo Experimental Forest, Puerto Rico". The final output is a graph depicting a 9 week average from the years between 1985 to 1995, from four locations in Puerto Rico: Bisley Quebrada 1-3 and Puente Roto Mameyes.

The figure displays a time-series plot for a 9 week moving average for the streamwater concentrations observed before and after the hurricane's impact on multiple watersheds and chemical ions. Thus, there is an initial increase during the year of the hurricane formation in 1989.

The importance of the paper was to highlight the lack of investigation on water chemistry during a watershed disturbance, specifically a hurricane, in tropical regions.

# Data

The data was received online through the Environmental Data Initiative portal created by William H. McDowell and the International Institute of Tropical Forestry(IITF), USDA Forest Service.

Figure 3 from Schaefer's article focused on the sites found in four separate data files. No other data was needed beyond those files.

```{r, echo = FALSE, include = FALSE}
library(tidyverse)
figure3 <- read_csv(
  "/Users/master/Downloads/Bren/MEDS Summer/eds214/MAH214final/output/clean_data.csv"
)
```


# Methods

The first step to reproducing the figure from Schaefer's article was to clean and combine the raw data by using the functions: select and filter.

The select function was used to choose the columns: the Sample Data, Sample ID, Nitrate, Potassium, Magnesium, Ammonium, and Calcium.

Then, filtering was done to sort through the correct years, 1988 to 1994, as the time period assigned for the window of observations which was then inputted to find the 9 week moving average.

This would then be graphed using ggplot by chemical ions and concentration levels dependent on site location.


# Results

The vertical line on the time-series plot is shown as the representation of the hurricane formation. Thus, the peaks are iniitally high while some ions may decrease post-hurricane, while others remain elevated.

Each ion, with the level plotted on the y-axis using each respectively found moving average across the 9 week window as time being depicted in years on the x-axis.

A legend shows the sites represented by dashed or solid lines for each ion.

The figure was created to show a close replica, though there have been some data loss over the years since its initial publication in 1999.

```{r, echo = FALSE}
ggplot(
  data = figure3, # reading the clean data csv in output
  mapping = aes(
    # gives the x-axis, y-axis, and the line by site location for each chemical ion
    x = Sample_Date,
    y = concentration,
    linetype = site
  )
) +
  ggtitle(
    "Hurricane effects on stream chemistry" # title from the article to give reference on what the figure is about
  ) +
  geom_line() +
  geom_vline(
    # to show reference line as shown in original figure which depicts the start of the distrubance
    xintercept = ymd("1989-09-18"), # hurricane hugo date formation
    linetype = "dashed", # dash line for the separation
    linewidth = 0.4
  ) +
  labs(
    x = "Years",
    y = NULL,
    caption = "Figure 3. Concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo, 9-wk moving averages."
  ) +
  facet_wrap(
    ~ion,
    ncol = 1,
    scales = "free",
    strip.position = "left",
    nrow = 5
  ) + # frees up the scaling for each ion since the units are differnet and so are the concentration levels
  theme_bw() + # black and white to match the original
  theme(
    legend.title = element_blank() # original did not have a legend title
  ) +
  theme(
    panel.grid.major = element_blank(), # remove major grid lines for cleaner look
    panel.grid.minor = element_blank() # remove minor grid lines for sleekness
  )
```

